### 2012-08-31 Dail_Mad_ext.tex
formatted intro

State-space models have become the most widespread approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_taper:1994,
  dennis_etal:2006}. Conventional state-space models are
simply time-series models in which the true state of the system is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow
\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let $N_t$ be
the abundance of a species during year $t$, and $X_t$ be the observed data,
which differs from  due to observation error ($\zeta$). A simple state
space model that includes exponential growth and process variation
($\grave{o}$) could be written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \\
X_t = N_t + \zeta_t
\end{gather}
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions:   and   (de
Valpine and Hastings 2002, Dennis et al. 2006). Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace   with   where A is the area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus . Generally, zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

The problems mentioned thus far are minor in comparison to the more
serious issue that the parameters of the model are often not
identifiable. Specifically, the Markovian nature of the model implies
the following likelihood
\[
LIKELIHOOD
\]
The first term is the probability density for the first observation in
the time series,  , yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation. This problem has resulted in several papers with
interesting names like ``multi-modal likelihoods and ridges etc...''
The workaround here is to assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.

@@

Dail and Madsen (2011) developed a model (henceforth the DM model)
that resolves each of these problems with traditional state-space
models. The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error using
only spatially and temporally replicated count data.  Their model is
an extension of the closed-population N-mixture model (Royle
2004)(2004), which was designed specifically to address the problem of
modeling spatial variation in abundance when detection probability is
less than unity. The DM extension relaxes the assumption of population
closure and includes explicit parameters describing population change
as a first-order Markovian process. [more details]

[applications / potential etc... ] The DM model is relevant to a huge
number of problems confronting basic and applied ecological
research. Answering questions about climate change etc... require
spatially and temporally extensive datasets such as result from the
North American Breeding Bird Survey (BBS) (Robbins et
al. 1986). However, most of the datasets of the appropriate scale are
collected by volunteers and using protocols that are not amenable to
traditional approaches of modeling abundance and detection probability
based on standard capture-recapture methods. Instead they have, at
best, simple count data for which, when it can be interpreted as
counts of unique individuals, the DM model can be applied.. Yadi ya

>>>>

State-space models have become the most widespread approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_taper:1994,
  dennis_etal:2006}. Conventional state-space models are
simply time-series models in which the true state of the system is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow
\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let $N_t$ be
the abundance of a species during year $t$, and $X_t$ be the observed data,
which differs from  due to observation error ($\zeta$). A simple state
space model that includes exponential growth and process variation
($\grave{o}$) could be written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \\
X_t = N_t + \zeta_t
\end{gather}
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions: $\grave{o}_t \sim
\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

The problems mentioned thus far are minor in comparison to the more
serious issue that the parameters of the model are often not
identifiable. Specifically, the Markovian nature of the model implies
the following likelihood
\[
LIKELIHOOD
\]
The first term is the probability density for the first observation in
the time series, $Y_t$, yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation. This problem has resulted in several papers with
interesting names like ``multi-modal likelihoods and ridges etc...''
The workaround here is to assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.

@@

\citet{dail_madsen:2011} developed a model (henceforth the DM model)
that resolves each of these problems with traditional state-space
models. The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error using
only spatially and temporally replicated count data.  Their model is
an extension of the closed-population N-mixture model
\citep{royle:2004biom}, which was designed specifically to address the
problem of
modeling spatial variation in abundance when detection probability is
less than unity. The DM extension relaxes the assumption of population
closure and includes explicit parameters describing population change
as a first-order Markovian process. [more details]

[applications / potential etc... ] The DM model is relevant to a huge
number of problems confronting basic and applied ecological
research. Answering questions about climate change etc... require
spatially and temporally extensive datasets such as result from the
North American Breeding Bird Survey (BBS)
\citep{robbins_etal:1986}. However, most of the datasets of the
appropriate scale are
collected by volunteers and using protocols that are not amenable to
traditional approaches of modeling abundance and detection probability
based on standard capture-recapture methods. Instead they have, at
best, simple count data for which, when it can be interpreted as
counts of unique individuals, the DM model can be applied.. Yadi ya
